Prediction of functional markers of mass cytometry data via deep learning

### 2019-02-07 revision/answers2reviewers.md

# Answers to reviewers

We thank the reviewers for such insightful comments which had certainly improved the manuscript greatly.

## From reviewer 1

1. "For the deep learning framework, there is not too much novelty. In my opinion, the paper is a pioneer of practitioner of the deep learning in their specific research area."
yes, we are not inventing neural networks, we are using them in this area where they have not been used before. we added this in the intro/discussion

2. "When they split the data into training, validation, test datasets, it’s better to follow the cross-validation schema, that is, try different repetition and report the average results to reduce the randomness."
definitely! we changed the analysis:  first the data was split into training set (1000,000 samples) and testing set (223,228 samples), and the training set was used to perform the 10-fold cross validation. Based on the average MSE from the cross validation, we chose the model with the smallest average MSE. Then we used all the training samples to obtain the final model and reported the prediction MSE of the testing samples.

3. "For the performance, we can see that the margin between NN and Random forest or linear model is not too much. SO, a better deep learning framework is still desired."
answer: in deep learning margins in performance can be relative. As we are comparing accuracy in prediction, random forest and NN are both quite competitive, with the former losing in computation time, rather than prediction accuracy. Other input is welcome. we added this in the intro/discussion

## From reviewer 2

1. "The study did not compare with support vector regression (SVR), citing “computational time constraints”. It is somewhat surprising since SVR is not a time-consuming algorithm comparing to deep learning.  Although the paper doesn’t have to do new work, some discussions along these lines could be added" 
It is our understanding that even when SVM are widely used in a variety of scenarios, SVR have not been as adopted in big data settings given its complexity.
One possible reason for the slow computation speed for SVR is if we are fitting a non-linear SVR then it could involve a radial basis function that involves computing a nXn kernel matrix where n is the number of samples which is very large for our application.

2. "The quality of the figures can be improved. Some labels in the figure are in light color, which is not easy to see. Some graphs (e.g. two in Fig. 2) have no axis labels."
Done, and figure 2 cannot have axis as it is the neural network architecture

3. "It would be good if the data used in this study can be made available to the public so that it can be served as a benchmark for studies in the community."
From Peng: The data is already publically available from the original paper that published the experiment, although one needs to learn a little bit about Cytobank to be able to get the data. Here is the link: http://reports.cytobank.org/1/v1

### 2019-02-07 revision/answers2reviewers.md

# Answers to reviewers

We thank the reviewers for such insightful comments which had certainly improved the manuscript greatly.

## From reviewer 1

1. "For the deep learning framework, there is not too much novelty. In my opinion, the paper is a pioneer of practitioner of the deep learning in their specific research area."
We agree! Neural networks are widely studied models that have been applied to a variety of real-life datasets. We are not proposing new model developments for this area of neural network, we are applying existing neural network tools to the case of flow cytometry area which has not fully adopted the deep learning schemes yet.
We added these points in the introduction/discussion

2. "When they split the data into training, validation, test datasets, it’s better to follow the cross-validation schema, that is, try different repetition and report the average results to reduce the randomness."
This is a great point! We have changed the analysis accordingly: first the data was split into training set (1000,000 samples) and testing set (223,228 samples), and the training set was used to perform the 10-fold cross validation. Based on the average MSE from the cross validation, we chose the model with the smallest average MSE. Then we used all the training samples to obtain the final model and reported the prediction MSE of the testing samples.

3. "For the performance, we can see that the margin between NN and Random forest or linear model is not too much. SO, a better deep learning framework is still desired."
We agree! Sometimes in deep learning comparisons, margins in performance can be relative. As we are comparing accuracy in prediction, random forest and NN are both quite competitive, with the former losing in computation time, rather than prediction accuracy. We added this discussion to the manuscript.

@@

## From reviewer 2

1. "The study did not compare with support vector regression (SVR), citing “computational time constraints”. It is somewhat surprising since SVR is not a time-consuming algorithm comparing to deep learning.  Although the paper doesn’t have to do new work, some discussions along these lines could be added" 
It is our understanding that even when SVM are widely used in a variety of scenarios, SVR models have not been successfully adopted in big data settings given their complexity.
One possible reason for the slow computation speed for SVR is that non-linear SVR involve a radial basis function that computes a n x n kernel matrix where n is the number of samples which is very large for our application.  We have added these details and some references to the paper.

2. "The quality of the figures can be improved. Some labels in the figure are in light color, which is not easy to see. Some graphs (e.g. two in Fig. 2) have no axis labels."
We have increased the font and changed the color of the figure labels. Figure 2 cannot have axis as it represents the neural network architecture.

3. "It would be good if the data used in this study can be made available to the public so that it can be served as a benchmark for studies in the community."
The data is already publically available from the original paper that published the experiment. Here is the link: http://reports.cytobank.org/1/v1
We added a section about Data Sharing in the manuscript.

### 2019-02-07 revision/latex_template/author.tex
modified manuscript according to reviews; only missing a few plots

Recently, deep learning models are revolutionizing the fields of
precision medicine, data mining, astronomy, human-computer
interactions, among many others, by becoming a major discovery force
in science due to the unprecedented accuracy in prediction.  Moreover,
deep learning approaches have shown accurate performance on genomics
and biomedical
applications\cite{Ciresan2013,Denas2013,Fakoor2013,Leung2014,CruzRoa2014,Li2017,Mobadersany2018}.

Furthermore, CyTOF data is perfectly suited for deep learning methods.
On one side, identify markers define a cell type (e.g., B cell, T
cell, monocytes, MSC), and on the other side, expressions of
functional markers identify the cell's activity (e.g., quiescent,
secreting cytokines, proliferating, apoptosis). Since CyTOF technology
allows for the simultaneous measurement of a large number of protein
markers, most CyTOF studies measure both identity markers and
functional markers, providing data for supervised learning tools, like
neural networks.  In addition, each CyTOF run typically collects data
on $10^6$ cells, creating an ideal large dataset in which the number
of samples (cells) is orders of magnitude larger than the number of
variables (markers). Deep learning methods are particularly suited for
this type of big data.

@@

Here, we explore neural network models to predict functional markers
(internal phosphoproteins) with identify markers (cell surface
proteins), and compare its performance in terms of accuracy and speed
to other standard statistical approaches like regression, and random
forests.  We show that neural networks improve prediction of
functional markers, making them a powerful alternative to the usual
regression techniques.

@@

We fit a neural network model to predict functional markers from
surface markers, and compare its performance to three classical
statistical methods: 1) linear regression (unpenalized and penalized),
2) decision trees, and 3) random trees. Due to computational time
constraints, we could not fit a support vector regression (SVR)
model. We compared the performance of the four approaches by computing
the mean square error (MSE) of the predicted responses.

To fit the models, we divided the data into training set, validation
set and testing set. The training set was used to fit each of the four
models. The validation set was used to determine the best setup
(tuning parameters) of each model in terms of MSE. Finally, the test
set was used to compare the MSE across methods.

The complete data consisted of 1,223,228 cells in 18 functional
markers (responses) and 15 surface markers (predictors), which we
divided as follows: 750,000 rows as training set, 250,000 rows as
validation set, and 223,228 rows as testing set.

@@

In this work, we showed that a neural network model outperforms
standard statistical approaches like linear regression and random
forest in the prediction of functional markers from surface markers
for CyTOF data. Neural networks were also faster and more efficient
than random forests, which make them a more viable choice for big
datasets.

The improved prediction accuracy of neural networks can be explained
by their flexibility to account for non-linearity or skewness.  Unlike
regression models, neural networks do not have linearity or normality
assumptions, and they take advantage of the correlation structure
among responses by fitting a network for the whole response vector.

@@

@@

>>>>

Recently, deep learning models are revolutionizing the fields of
precision medicine, data mining, astronomy, human-computer
interactions, among many others, by becoming a major discovery force
in science due to the unprecedented accuracy in prediction.
\claudia{Moreover}{However}, \claudia{}{despite} deep learning
approaches \claudia{have shown}{showing} accurate performance on
genomics and biomedical
applications\cite{Ciresan2013,Denas2013,Fakoor2013,Leung2014,CruzRoa2014,Li2017,Mobadersany2018},
\claudia{}{the CyTOF community has not fully adopted these
  approaches.}  \claudia{Furthermore}{In fact, it turns out that}
CyTOF data is perfectly suited for deep learning methods.  On one
side, identify markers define a cell type (e.g., B cell, T cell,
monocytes, MSC), and on the other side, expressions of functional
markers identify the cell's activity (e.g., quiescent, secreting
cytokines, proliferating, apoptosis). Since CyTOF technology allows
for the simultaneous measurement of a large number of protein markers,
most CyTOF studies measure both identity markers and functional
markers, providing data for supervised learning tools, like neural
networks.  In addition, each CyTOF run typically collects data on
$10^6$ cells, creating an ideal large dataset in which the number of
samples (cells) is orders of magnitude larger than the number of
variables (markers). Deep learning methods are particularly suited for
this type of big data.

@@

Here, we \claudia{}{present the first work to} explore neural network
models to predict functional markers (internal phosphoproteins) with
identify markers (cell surface proteins). We compare the performance
of neural networks in terms of accuracy and speed to other standard
statistical approaches like regression, and random forests.  We show
that neural networks improve prediction of functional markers, making
them a powerful alternative to the usual regression techniques,
\claudia{}{thus, providing quantitive evidence that deep learning
  methods can enrich the existing research landscape of CyTOF big
  data.}

@@

We fit a neural network model to predict functional markers from
surface markers, and compare its performance to three classical
statistical methods: 1) linear regression (unpenalized and penalized),
2) decision trees, and 3) random trees. Due to computational time
constraints, we could not fit a support vector regression (SVR) model
(more in Discussion). We compared the performance of the four
approaches by computing the mean square error (MSE) of the predicted
responses.

To fit the models, we divided the data into training set,
\claudia{validation set}{} and testing set. The training set was used
to \claudia{fit each of the four models}{perform a 10-fold cross
  validation scheme to choose the best model fit for each method. We
  chose the model fit with the smallest average MSE}. \claudia{The
  validation set was used to determine the best setup (tuning
  parameters) of each model in terms of MSE. Finally, the test set was
  used to compare the MSE across methods.}{Then, we used all the
  training samples to estimate the final model for each method
  (e.g. model and tuning parameters), and reported the prediction MSE
  of each method based on the testing samples.}

The complete data consisted of 1,223,228 cells in 18 functional
markers (responses) and 15 surface markers (predictors), which we
divided as follows: \claudia{750,000}{ 1,000,000} rows as training set,
\claudia{250,000 rows as validation set}{}, and 223,228 rows as
testing set.

@@

\claudia{}{To the best of our knowledge, our work presents the first
  neural network model for the prediction of functionality in CyTOF
  big data.}  In this work, we showed that a neural network model
outperforms standard statistical approaches like linear regression and
random forest in the prediction of functional markers from surface
markers for CyTOF data. Neural networks were also faster and more
efficient than random forests, which make them a more viable choice
for big datasets.

The improved prediction accuracy of neural networks can be explained
by their flexibility to account for non-linearity or skewness.  Unlike
regression models, neural networks do not have linearity or normality
assumptions, and they take advantage of the correlation structure
among responses by fitting a network for the whole response vector.
\claudia{}{It is important to note that the while the neural network
  model shows better performance than the other methods, this edge is
  still quite modest. Random forest models also represent a suitable
  choice for prediction under non-linear assumptions. However, we show
  here that when scalability also matters, the neural network model
  outperforms random forests (in computation time), while providing
  more accurate predictions. We claim that neural networks are better
  suited for big CyTOF data than random forests, but more work is
  needed to corroborate this claims.}

@@

\claudia{}{Initially, we were surprised by the lack of scalability of
  SVR models. However, given the non-linearity and optimization
  complexity\cite{Smola2003, Welling2004}, it appeared that our
  dataset was too large for the computation of the kernel matrix,
  which is $n \times n$ where $n$ is the number of samples (more than
  1 million in our case). More work is needed to assess the
  performance of SVR under smaller datasets to assess the accuracy of
  such models and the limits in scalability.}

@@

\vspace{0.5cm}
\noindent \textbf{Acknowledgements:} We thank the editor as well as
two anonymous reviewers whose suggestions greatly improved the
manuscript.

\vspace{0.5cm}
\noindent \textbf{Data Sharing:} The data is publicly available
through the original publications\cite{Bendall2011,Qiu2011} in the
following website: \url{http://reports.cytobank.org/1/v1}.
